Appendix to "Fast approximate Bayesian inference of HIV indicators using PCA adaptive Gauss-Hermite quadrature

### 2023-07-20 src/dev_scale-grid/scale-grid.Rmd
Add hyperparameter names to PC loadings plot, add variance explained number (not just near 90), work on formatting of tables #55

tv <- cumsum(eigenC$values) / sum(eigenC$values)
```

@@

```{r pc-loadings, class.source = 'fold-hide', fig.cap="PC loadings."}
rownames(E) <- hypers
plot_pc_loadings(eigenC)

>>>>

tv <- cumsum(eigenC$values) / sum(eigenC$values)

tv_df <- data.frame(
  s = 1:length(m),
  tv = tv
)

write_csv(tv_df, "total-variation.csv")
```

@@

```{r pc-loadings, class.source = 'fold-hide', fig.cap="PC loadings."}
rownames(eigenC$vectors) <- hypers
plot_pc_loadings(eigenC)

### 2023-07-20 src/docs_paper/appendix.Rmd
Add hyperparameter names to PC loadings plot, add variance explained number (not just near 90), work on formatting of tables #55

In this section we describe the simplified Naomi model [@eaton2021naomi] in more complete mathematical detail than the main text.

@@

Consider the most recent national household survey with HIV testing which has taken place in the country of interest.
Let $x \in \{1, \ldots, n\}$ refer to a district located within the Spectrum [@stover2010spectrum] region $R_x$.
Let $s \in \{\text{F}, \text{M}\}$ be sex, and $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, \text{80+}\}$ be five-year age groups.
As short-hand, we write $a = l$ to refer to the age group with lower bound $l$, e.g. $a = 20$ for $a = \text{20-25}$.

@@

Sometimes data are observed at an aggregate level, rather than the more granular modelled level.
We use $\{\cdot\}$ to generically refer to a aggregate set over which an observation is made, e.g. $\{a\} = \{\text{15-19}, \ldots, \text{45-49}\}$ for the adult age range 15-49.
The notation $\sum_{\{x\}}$ is used as shorthand for $\sum_{x \in \{x\}}$, and analogously for $s$ and $a$.

@@

1. We specify the first order auto-regressive model by $\bu \sim \text{AR}1(\sigma, \phi)$ such that
\begin{align}
u_1 &\sim \left( 0, \frac{1}{1 - \rho^2} \right), \\
u_i &= \rho u_{i - 1} + \epsilon_i, \quad i = 2, \ldots, \dim(\bu)
\end{align}
where $\epsilon_i \sim \mathcal{N}(0, 1)$ are independent and identically distributed (IID) Gaussian random variables, $\sigma > 0$ is the marginal standard deviation, and $|\rho| < 1$ is the (lag-one) correlation parameter.
2. We use $\bu \sim \text{ICAR}(\sigma)$ to refer to the Besag intrinsic conditional auto-regressive model (ICAR) [@besag1991bayesian] with full conditionals
\begin{equation}
u_i \, | \, \bu_{-i} \sim \mathcal{N} \left(\frac{\sum_{j: j \sim i} u_j}{n_{\delta i}}, \frac{\sigma^2}{n_{\delta i}}\right),
\end{equation}
where $\bu_{-i}$ is $\bu$ with the $i$th element removed i.e. $(u_1, \ldots, u_{i - 1}, u_{i + 1}, \ldots, u_{\dim(\bu)})$, $j \sim i$ if the units $i$ and $j$ are defined as adjacent, $n_{\delta i} = |\{j:j \sim i\}|$ is the total number of adjacent units, and $\sigma > 0$ is the marginal standard deviation.
We follow recommendations of @freni2018note on scaling of precision matrices, disconnected adjacency graph components, and islands.
3. For the reparameterised Besag-York-Mollie model (BYM2) [@simpson2017penalising], we write $\bu \sim \text{BYM}2(\sigma, \phi)$, where $\bu$ is comprised of a spatially structured ICAR component $\bv$ with proportion $\phi \in (0, 1)$ and spatially unstructured IID component $\w$ with proportion $1 - \phi$, both scaled to have generalised variance equal to one, and $\sigma > 0$ is the marginal standard deviation such that
\begin{equation}
\bu = \sigma \left( \sqrt\phi \cdot \bv + \sqrt{1 - \phi} \cdot \w \right).
\end{equation}

@@

We now describe the hyperparameter and latent field process specification for the model.
Whereas in the main text process and likelihood specifications are written together, here we consider the likelihood equations separately in Section \ref{sec:likelihood} to follow.
Table \ref{tab:latent-hyper-size} gives the number of latent field parameters and hyperparameters in each component of the model.
In the case of Malawi then $n = 32$ such that the total number of latent field parameters is $51 + 14 \cdot 32 = 499$ and the total number of hyperparameters is $24$.

@@

```{r scree, fig.cap="Most variation could be explained by far fewer than the full 24 dimensions."}
knitr::include_graphics("depends/total-variation.png")
```

```{r reduced-rank, fig.cap="The reduced rank (8) matrix approximation to the covariance matrix is visually similar to the full rank original."}
knitr::include_graphics("depends/reduced-rank.png")
```

@@

```{r pc-loadings, fig.cap="Principal component loadings."}
knitr::include_graphics("depends/pc-loadings.png")
```

@@

ks_summary %>%
  ungroup() %>%
  filter(type == "Latent field") %>%
  select(-type) %>%
  rename(`PCA-AGHQ` = `KS(aghq, tmbstan)`, `TMB` = `KS(TMB, tmbstan)`) %>%
  gt::gt(rowname_col = "Parameter") %>%
  gt::fmt_number(
    columns = c(`PCA-AGHQ`, `TMB`),
    decimals = 3
  ) %>%
  gt::summary_rows(
   columns = everything(),
   fns = list(Average = ~ mean(.x)),
   formatter = gt::fmt_number,
   decimals = 3
  )
```

>>>>

In this section we describe the simplified Naomi model [@eaton2021naomi] in complete detail.

@@

Consider the most recent national household survey with HIV testing which has taken place in the country of interest.
Let $x \in \{1, \ldots, n\}$ be district, $s \in \{\text{F}, \text{M}\}$ be sex, and $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, \text{80+}\}$ be five-year age groups.
Each district is located within a correspodning Spectrum [@stover2010spectrum] region $R_x$.
As short-hand, we write $a = l$ to refer to the age group with lower bound $l$, e.g. $a = 20$ for $a = \text{20-25}$.

@@

Sometimes data are observed at an aggregate level, rather than the more granular modelled level.
We use $\{\cdot\}$ to generically refer to a aggregate set over which an observation is made, e.g. $\{a\} = \{\text{15-19}, \ldots, \text{45-49}\}$ for the adult age range 15-49.
The notation $\sum_{\{x\}}$ is used as shorthand for $\sum_{x \in \{x\}}$.
Analogous notion is used for sums over $\sum_{a \in \{a\}}$ and $\sum_{s \in \{s\}}$.

@@

1. We specify the first order auto-regressive model by $\bu \sim \text{AR}1(\sigma, \phi)$ such that
\begin{align}
u_1 &\sim \left( 0, \frac{1}{1 - \rho^2} \right), \\
u_i &= \rho u_{i - 1} + \epsilon_i, \quad i = 2, \ldots, \dim(\bu),
\end{align}
where $\epsilon_i \sim \mathcal{N}(0, 1)$ are independent and identically distributed (IID) Gaussian random variables, $\sigma > 0$ is the marginal standard deviation, and $|\rho| < 1$ is the (lag-one) correlation parameter.
2. We use $\bu \sim \text{ICAR}(\sigma)$ to refer to the Besag intrinsic conditional auto-regressive model (ICAR) [@besag1991bayesian] with full conditionals
\begin{equation}
u_i \, | \, \bu_{-i} \sim \mathcal{N} \left(\frac{\sum_{j: j \sim i} u_j}{n_{\delta i}}, \frac{\sigma^2}{n_{\delta i}}\right),
\end{equation}
where $\bu_{-i}$ is $\bu$ with the $i$th element removed i.e. $(u_1, \ldots, u_{i - 1}, u_{i + 1}, \ldots, u_{\dim(\bu)})$, $j \sim i$ if the units $i$ and $j$ are defined as adjacent, $n_{\delta i} = |\{j:j \sim i\}|$ is the total number of adjacent units, and $\sigma > 0$ is the marginal standard deviation.
We follow recommendations of @freni2018note on scaling of precision matrices, disconnected adjacency graph components, and islands.
3. For the reparameterised Besag-York-Mollie model [BYM2; @simpson2017penalising], we write $\bu \sim \text{BYM}2(\sigma, \phi)$, where $\bu$ is comprised of a spatially structured ICAR component $\bv$ with proportion $\phi \in (0, 1)$ and spatially unstructured IID component $\w$ with proportion $1 - \phi$, both scaled to have generalised variance equal to one, and $\sigma > 0$ is the marginal standard deviation such that
\begin{equation}
\bu = \sigma \left( \sqrt\phi \cdot \bv + \sqrt{1 - \phi} \cdot \w \right).
\end{equation}

@@

We now describe the hyperparameter and latent field process specification for the model.
Whereas in the main text process and likelihood specifications are given together, here we consider the likelihood equations separately (Section \ref{sec:likelihood}).
Table \ref{tab:latent-hyper-size} gives the number of latent field parameters and hyperparameters in each component of the model.
In the case of Malawi then $n = 32$ such that the total number of latent field parameters is $51 + 14 \cdot 32 = 499$ and the total number of hyperparameters is $24$.

@@

```{r scree, fig.cap="Most variation in the inverse curvature could be explained using far fewer than the full 24 dimensions."}
knitr::include_graphics("depends/total-variation.png")
```

```{r reduced-rank, fig.cap="The reduced rank ($s = 8$) matrix approximation to the covariance matrix is visually similar to the full rank original."}
knitr::include_graphics("depends/reduced-rank.png")
```

@@

```{r pc-loadings, fig.cap="Principal component loadings based on the eigendecomposition of the inverse curvature."}
knitr::include_graphics("depends/pc-loadings.png")
```

@@

ks_summary %>%
  ungroup() %>%
  filter(type == "Latent field") %>%
  select(-type) %>%
  rename(TMB = `KS(TMB, tmbstan)`, `PCA-AGHQ` = `KS(aghq, tmbstan)`) %>%
  select(Parameter, TMB, `PCA-AGHQ`) %>%
  mutate(Difference = TMB - `PCA-AGHQ`) %>%
  gt::gt() %>%
  gt::summary_rows(
   columns = c(`TMB`, `PCA-AGHQ`, Difference),
   fns = list(Average = ~ mean(.x)),
   formatter = gt::fmt_number,
   decimals = 3
  ) %>%
  gt::fmt_number(
    columns = c(`TMB`, `PCA-AGHQ`, Difference),
    decimals = 3
  ) %>%
  gt::text_transform(
    locations = gt::cells_body("Parameter"),
    fn = function(x) paste0("\\texttt{", x, "}")
  ) %>%
  gt::text_transform(
    locations = gt::cells_body("Difference"),
    fn = function(x) ifelse(x > 0, paste0("\\textbf{", x, "}"), paste0(x))
  ) %>%
  gt::as_latex() %>%
  as.character() %>%
  cat(file = "ks-summary.tex")
```

\input{ks-summary.tex}
